Verifying synchronous distributed algorithms using IVy

### 2018-08-12 paper/paper.tex
- page 13, section (IV.iii): don't -> do not, incomming_message -> 	incoming_message, hasn't -> has not

\textit{Assume} is used to exclude the possibility of something happening.  This is mostly useful in combination with non-deterministic assignments that one wants to restrict. The following action chooses a random node to assign to \textit{y}, but will not allow control to pass if the chosen node has failed. Therefore the actions will choose a random non-failed node to send a message to x.  If there is no node that hasn't failed, this action cannot terminate.

@@

\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
action round = {
  var iter := node.t.iter.create(0);
  while ~iter.is_end
  {
    call send_messages(iter.val)
    iter := iter.next;
  }
  . . .
}
action send_messages(n:node.t) = {
  node.message(n,V) := node.w(n,V);
  node.incomming_message(O,n) := true;
}

@@

\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
. . .
line 3: message := *;
line 3: incomming_message := *;
. . .
\end{lstlisting}
\end{mdframed}

IVy non-deterministically assigns to all relations involved in the loop, instead of actually executing the loop.  This is because proving correctness of loops can only be done through inductive loop invariants specified by a human. IVy therefore allows specifying invariants together with a loop.
\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
  while ~iter.is_end
  invariant node.message(N,V) = node.w(N,V)
  invariant N <= iter.val -> incomming_message(O,N)
  {
    . . .
  }
\end{lstlisting}
\end{mdframed}

@@

\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
assert message(N,V) = w(N,V)
assert N:t <= t.iter.val(loc:iter) -> incomming_message(O,N)
incomming_message := *
message := *
assume message(N,V) = w(N,V)
assume N:t <= t.iter.val(loc:iter) -> incomming_message(O,N)
\end{lstlisting}
\end{mdframed}

@@

  # incomming_message(n,o) denotes whether node n received a message
  # from node o
  relation incomming_message(SELF:node, OTHER:node)

@@

  # helper to simulate non-determinstic message sending
  relation incomming_message_failing(SELF:node, OTHER:node)

@@

\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
after init {
  def := *;
  default(N) := def;
  w(N, V) := V=node.val(N);
  incomming_message(A, B) := false;
  decision(N, V) := false;
  clean_round_occured := false;
  decision_round := false;
  node.failed_before_round(N) := false;
  node.failed(N) := false;
  node.has_single_value(N) := false;
}
\end{lstlisting}
\end{mdframed}

@@

\subsection{Step 1 - Message sending}
The contents of a node's message is modeled via the relation $message(N:node,V:value)$. Where
$message(1,0) = true; message(1,1) = true;$ means node $1$ includes in its message the values $0$ and $1$.
Since a node sends the same message to all nodes, for any given round, we don't need to include the
recipient in the relation.

The actual delivery of a message is modeled via the $incomming\_message(RECIPIENT:node,SENDER:node)$ relation. Where $incoming\_message(0,1) = true; incoming\_message(0,2) = true;$ means that node $0$ receives the messages from node $1$ and node $2$.

These relations allow us to use IVy's assignments to easily model \textit{Step 1} in a round of \textit{Floodset}.
Line 2 assigns for all nodes $A$ and values $V$ to $message(A,V)$ the boolean value of $node.w(A,V) \land \neg failed(A)$. In other words, every node that hasn't failed prepares a message with all values in its $w$. Line 3 states that all nodes that haven't failed send messages to all other nodes.

\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
action step1 = {
  node.message(A,V) := node.w(A,V) & ~node.failed(A);
  node.incomming_message(O,A) := ~node.failed(A);
}
\end{lstlisting}
\end{mdframed}

@@

Line 2 in the \textit{step2} action adds to $w$ of node $N$ all values $V$ for which a node $J$ exists that has sent a message to $N$ that contains $J$.
The conditional statement in Line 4 corresponds to the \textit{decision condition} ($if\ rounds = f+1$) in the original algorithm. Nodes with a single value in $w$ decide on that value, otherwise
they decide on the default value.
Finally, for good measure, $incomming\_message$ and $message$ are reset, but this is not necessary for the proof.

\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
action step2 = {
  node.w(N,V) := (node.w(N,V) |
                 (~node.failed(N) &
                  exists J . (node.incomming_message(N,J) &
                               node.message(J,V))));

@@

  node.incomming_message(A,B) := false;
  node.message(A,V) := false;
}
\end{lstlisting}
\end{mdframed}

@@

To recap, a process may fail before or after performing some instance of Step 1 or Step 2 as well as in the middle of performing Step 1.  The presented code for actions \textit{step1} and \textit{step2} already correctly considers the behaviour of nodes that failed before the actions. E.g. \textit{step1} only allows nodes that have not failed to send messages.
\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
. . .
node.incomming_message(O,A) := ~node.failed(A);
. . .
\end{lstlisting}
\end{mdframed}

@@

\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
node.incomming_message_failing(N,M) := *;
node.incomming_message(N,M) :=
  (node.failed_during_round(M) &
    node.incomming_message_failing(N,M)) |
  (~node.failed_during_round(M) &
    node.incomming_message(N,M));
\end{lstlisting}
\end{mdframed}

Line 1 prepares for each node $M$ a non-deterministically chosen set of nodes that it sends a message to. This set is represented by the relation \textit{incomming\_message\_failing}. This relation can be thought of as the set of nodes $M$ will send a message to if $M$ fails.
Line 2 updates the \textit{incomming\_message} relation. If node $M$ failed during this round, the set of nodes
 $M$ sends messages to is changed to the values of \textit{incomming\_message\_failing}, otherwise it remains the same. It is important to point out that this approach correctly simulates the fact that nodes are not thought of as sending messages sequentially but in non-deterministic order. The specification makes no assumptions about an order of message sending. Furthermore, for fixed $m$, \textit{incomming\_message\_failing(N,m)} may still be $true$ for all N. The specification therefore includes the possibility of a node failing after it sent all its messages successfully.

@@

    node.incomming_message_failing(N,M) := *;
    node.incomming_message(N,M) :=
      (node.failed_during_round(M) &
        node.incomming_message_failing(N,M)) |
      (~node.failed_during_round(M) &
        node.incomming_message(N,M));

@@

# 'after' a node has failed it will not send anything
invariant node.failed_before_round(N) -> ~node.incomming_message(M, N)

@@

\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
node.w(N,V) := (node.w(N,V) |
               (~node.failed(N) &
                exists J . (node.incomming_message(N,J) &
                             node.message(J,V))));
\end{lstlisting}
\end{mdframed}

@@

\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
node.tmp_w(A,V) := node.w(A V);
node.w(A,V) := *;
assume node.w(A,V) <-> (node.tmp_w(A,V) |
                        ((node.incomming_message(A,J) &
                          node.message(J,V))));
\end{lstlisting}
\end{mdframed}
The second snippet misses an existential quantifier on line 4, and therefore specifies behaviour that is unlike that of \textit{Floodset}.  The author did not discover that this $assume$ is bogus and excludes valid system states from being verified. The error only became apparent when it was discovered that correctness could be proven in the absence of an invariant that should have been absolutely necessary for induction.

@@

\noindent For example, one may insert
\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
assert ~(exists N . node.failed(N) &
        (exists J . ~incoming_message(J,N) &
          exists K . incomming_message(K,N) & J~=N & K~=J & N~=K))
\end{lstlisting}
\end{mdframed}
after the \textit{simulate\_failure} action. The negated part, $\gamma$, specifies that there exists a node $N$, that failed and failed to deliver a message to $J$ while successfully delivering a message to $K$, with $N,J,K$ distinct. By negating $\gamma$, we force Ivy to construct a counterexample such that $\gamma$ is satisfied. If IVy does not find a counterexample to this assertion, it means that $\gamma$ cannot occur. In this particular case, that would mean the specification does not allow partial message delivery. The specification would therefore be erroneous. One has to finally remove the failing assertion to allow Ivy to prove induction again.

>>>>

\textit{Assume} is used to exclude the possibility of something happening.  This is mostly useful in combination with non-deterministic assignments that one wants to restrict. The following action chooses a random node to assign to \textit{y}, but will not allow control to pass if the chosen node has failed. Therefore the actions will choose a random non-failed node to send a message to x.  If there is no node that has not failed, this action cannot terminate.

@@

\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
action round = {
  var iter := node.t.iter.create(0);
  while ~iter.is_end
  {
    call send_messages(iter.val)
    iter := iter.next;
  }
  . . .
}
action send_messages(n:node.t) = {
  node.message(n,V) := node.w(n,V);
  node.incoming_message(O,n) := true;
}

@@

\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
. . .
line 3: message := *;
line 3: incoming_message := *;
. . .
\end{lstlisting}
\end{mdframed}

IVy non-deterministically assigns to all relations involved in the loop, instead of actually executing the loop.  This is because proving correctness of loops can only be done through inductive loop invariants specified by a human. IVy therefore allows specifying invariants together with a loop.
\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
  while ~iter.is_end
  invariant node.message(N,V) = node.w(N,V)
  invariant N <= iter.val -> incoming_message(O,N)
  {
    . . .
  }
\end{lstlisting}
\end{mdframed}

@@

\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
assert message(N,V) = w(N,V)
assert N:t <= t.iter.val(loc:iter) -> incoming_message(O,N)
incoming_message := *
message := *
assume message(N,V) = w(N,V)
assume N:t <= t.iter.val(loc:iter) -> incoming_message(O,N)
\end{lstlisting}
\end{mdframed}

@@

  # incoming_message(n,o) denotes whether node n received a message
  # from node o
  relation incoming_message(SELF:node, OTHER:node)

@@

  # helper to simulate non-determinstic message sending
  relation incoming_message_failing(SELF:node, OTHER:node)

@@

\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
after init {
  def := *;
  default(N) := def;
  w(N, V) := V=node.val(N);
  incoming_message(A, B) := false;
  decision(N, V) := false;
  clean_round_occured := false;
  decision_round := false;
  node.failed_before_round(N) := false;
  node.failed(N) := false;
  node.has_single_value(N) := false;
}
\end{lstlisting}
\end{mdframed}

@@

\subsection{Step 1 - Message sending}
The contents of a node's message is modeled via the relation $message(N:node,V:value)$. Where
$message(1,0) = true; message(1,1) = true;$ means node $1$ includes in its message the values $0$ and $1$.
Since a node sends the same message to all nodes, for any given round, we do not need to include the
recipient in the relation.

The actual delivery of a message is modeled via the $incoming\_message(RECIPIENT:node,SENDER:node)$ relation. Where $incoming\_message(0,1) = true; incoming\_message(0,2) = true;$ means that node $0$ receives the messages from node $1$ and node $2$.

These relations allow us to use IVy's assignments to easily model \textit{Step 1} in a round of \textit{Floodset}.
Line 2 assigns for all nodes $A$ and values $V$ to $message(A,V)$ the boolean value of $node.w(A,V) \land \neg failed(A)$. In other words, every node that has not failed prepares a message with all values in its $w$. Line 3 states that all nodes that haven't failed send messages to all other nodes.

\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
action step1 = {
  node.message(A,V) := node.w(A,V) & ~node.failed(A);
  node.incoming_message(O,A) := ~node.failed(A);
}
\end{lstlisting}
\end{mdframed}

@@

Line 2 in the \textit{step2} action adds to $w$ of node $N$ all values $V$ for which a node $J$ exists that has sent a message to $N$ that contains $J$.
The conditional statement in Line 4 corresponds to the \textit{decision condition} ($if\ rounds = f+1$) in the original algorithm. Nodes with a single value in $w$ decide on that value, otherwise
they decide on the default value.
Finally, for good measure, $incoming\_message$ and $message$ are reset, but this is not necessary for the proof.

\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
action step2 = {
  node.w(N,V) := (node.w(N,V) |
                 (~node.failed(N) &
                  exists J . (node.incoming_message(N,J) &
                               node.message(J,V))));

@@

  node.incoming_message(A,B) := false;
  node.message(A,V) := false;
}
\end{lstlisting}
\end{mdframed}

@@

To recap, a process may fail before or after performing some instance of Step 1 or Step 2 as well as in the middle of performing Step 1.  The presented code for actions \textit{step1} and \textit{step2} already correctly considers the behaviour of nodes that failed before the actions. E.g. \textit{step1} only allows nodes that have not failed to send messages.
\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
. . .
node.incoming_message(O,A) := ~node.failed(A);
. . .
\end{lstlisting}
\end{mdframed}

@@

\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
node.incoming_message_failing(N,M) := *;
node.incoming_message(N,M) :=
  (node.failed_during_round(M) &
    node.incoming_message_failing(N,M)) |
  (~node.failed_during_round(M) &
    node.incoming_message(N,M));
\end{lstlisting}
\end{mdframed}

Line 1 prepares for each node $M$ a non-deterministically chosen set of nodes that it sends a message to. This set is represented by the relation \textit{incoming\_message\_failing}. This relation can be thought of as the set of nodes $M$ will send a message to if $M$ fails.
Line 2 updates the \textit{incoming\_message} relation. If node $M$ failed during this round, the set of nodes
 $M$ sends messages to is changed to the values of \textit{incoming\_message\_failing}, otherwise it remains the same. It is important to point out that this approach correctly simulates the fact that nodes are not thought of as sending messages sequentially but in non-deterministic order. The specification makes no assumptions about an order of message sending. Furthermore, for fixed $m$, \textit{incoming\_message\_failing(N,m)} may still be $true$ for all N. The specification therefore includes the possibility of a node failing after it sent all its messages successfully.

@@

    node.incoming_message_failing(N,M) := *;
    node.incoming_message(N,M) :=
      (node.failed_during_round(M) &
        node.incoming_message_failing(N,M)) |
      (~node.failed_during_round(M) &
        node.incoming_message(N,M));

@@

# 'after' a node has failed it will not send anything
invariant node.failed_before_round(N) -> ~node.incoming_message(M, N)

@@

\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
node.w(N,V) := (node.w(N,V) |
               (~node.failed(N) &
                exists J . (node.incoming_message(N,J) &
                             node.message(J,V))));
\end{lstlisting}
\end{mdframed}

@@

\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
node.tmp_w(A,V) := node.w(A V);
node.w(A,V) := *;
assume node.w(A,V) <-> (node.tmp_w(A,V) |
                        ((node.incoming_message(A,J) &
                          node.message(J,V))));
\end{lstlisting}
\end{mdframed}
The second snippet misses an existential quantifier on line 4, and therefore specifies behaviour that is unlike that of \textit{Floodset}.  The author did not discover that this $assume$ is bogus and excludes valid system states from being verified. The error only became apparent when it was discovered that correctness could be proven in the absence of an invariant that should have been absolutely necessary for induction.

@@

\noindent For example, one may insert
\begin{mdframed}[nobreak=true, backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
assert ~(exists N . node.failed(N) &
        (exists J . ~incoming_message(J,N) &
          exists K . incoming_message(K,N) & J~=N & K~=J & N~=K))
\end{lstlisting}
\end{mdframed}
after the \textit{simulate\_failure} action. The negated part, $\gamma$, specifies that there exists a node $N$, that failed and failed to deliver a message to $J$ while successfully delivering a message to $K$, with $N,J,K$ distinct. By negating $\gamma$, we force Ivy to construct a counterexample such that $\gamma$ is satisfied. If IVy does not find a counterexample to this assertion, it means that $\gamma$ cannot occur. In this particular case, that would mean the specification does not allow partial message delivery. The specification would therefore be erroneous. One has to finally remove the failing assertion to allow Ivy to prove induction again.
